---
title: "Facets Home"
author: "ESWebbink"
date: "8/3/2025"
---


```{r general, figures-side, out.width="50%", fig.height=3.0, message=FALSE, echo=FALSE}

# load libraries ####
library(tidyverse)
library(plotly)
library(readr)
library(knitr)

```
*** 
#### **Real Gross Domestic Product is the output of an economy adjusted for inflation in prices**
##### Real GDP is shown annualized, but on a non seasonally adjusted basis given the unusual impact of CoVid.<br> Potential GDP is a smoother, seasonally adjusted estimate based on projected growth in productivity and working age population - a projection unlikely to have taken CoVid's impact into account as yet.
<!-- | | -->
<!-- | :-: | -->
<!-- | *Real GDP in Billions of Chained 2012 Dollars at Annual Rates* | -->
```{r gdp_plot, figures-side, out.width="100%", fig.height=3, message=FALSE, echo=FALSE, warning=FALSE}

# read in quarterly data ####
df_q_gdp <- read_tsv("data/Facets_Quarterly_API.txt")
# colnames(df_q_gdp)

# change series names to make them more readable
colnames(df_q_gdp) <- c("Yr_Mo", "Real_GDP_Ch_vs_Yr_Ago", "Real_GDP_Ch_vs_Prior_Qtr", "Population", "Capacity_Utilization", "Real_Potential_GDP", "Real_GDP")
# colnames(df_q_gdp)

# Drop all rows where [real/not potential] GDP = NA
df_q_gdp <- df_q_gdp[is.na(df_q_gdp$Real_GDP)==FALSE,]

# transform data so both series are in annual terms
df_q_gdp$Real_GDP_Annualized <- 4 * df_q_gdp$Real_GDP

# subset GDP Real & Potential $Billions, then flip the frame - to draw a 2-line chart
df_q_billion <- df_q_gdp[,c("Yr_Mo", "Real_Potential_GDP","Real_GDP_Annualized")]
df_q_flip <- gather(df_q_billion, 
                     key = "GDP", 
                     value = "Bill_2012_Chain", 
                     -Yr_Mo,
                     na.rm=FALSE)

# plot the two variables on one line chart
p1 <- ggplotly(
  dynamicTicks = TRUE, 
  originalData = FALSE,
  ggplot(data = df_q_flip, 
         aes(x=Yr_Mo, 
             color=GDP, 
             y=Bill_2012_Chain)) +
    geom_line() + 
    scale_colour_hue(l = 40, 
                     h = c(160, 300) + 15,
                     c = 100) +
    theme(panel.background = element_rect(fill = "#eeeeee", colour = "grey"),
          text = element_text(size=8),
          legend.position = "left") +
    labs(y="Billions of Dollars"),
         title="Real GDP in Billions of Chained 2012 Dollars at Annual Rates")
#    facet_wrap(~ GDP, nrow = 1)
# )

# var layout = {
#   height: 230,
#   width: 100,
#   margin: {
#     l: 60,
#     r: 10,
#     b: 0,
#     t: 10,
#     pad: 4
#   }
# };

# Plotly.plot(p1, pieData, layout);

p1
# config(p1, displayModeBar = FALSE)
 # layout(p1, margin = list(t = 0, r = 0, b = 0, l = 0, unit = "pt", pad = 0))

```

<!-- | | | | -->
<!-- | :-: | :-: | :-: | -->
<!-- | *Growth over Prior Qtr* | *Growth over Prior Yr* | *Capacity Utilization* | -->

```{r gdpch_plots, figures-side, out.width="50%", fig.height=5, message=FALSE, echo=FALSE}

# q plots ####
  
  ggplot(data = df_q_gdp, aes(x = Yr_Mo, y = Real_GDP_Ch_vs_Prior_Qtr)) +
    geom_line(colour = "darkorchid") +
    theme(panel.background = element_rect(fill = "#eeeeee"), 
          text = element_text(size = 18)) +
    labs(y="Percent",
         x="Year/Qtr",
         title="Growth over Prior Qtr") +
    ylim(min = min(df_q_gdp$Real_GDP_Ch_vs_Yr_Ago, na.rm = TRUE),
         max = max(df_q_gdp$Real_GDP_Ch_vs_Yr_Ago, na.rm = TRUE))
  
  ggplot(data = df_q_gdp, aes(x = Yr_Mo, y = Real_GDP_Ch_vs_Yr_Ago)) +
    geom_line(colour = "darkorchid") +
    theme(panel.background = element_rect(fill = "#eeeeee"),
                    text = element_text(size = 18)) +
    labs(y="Percent",
         x="Year/Qtr",
         title="Growth over Prior Yr") +
    ylim(min = min(df_q_gdp$Real_GDP_Ch_vs_Yr_Ago, na.rm = TRUE),
         max = max(df_q_gdp$Real_GDP_Ch_vs_Yr_Ago, na.rm = TRUE))

  ggplot(data = df_q_gdp, aes(x = Yr_Mo, y = Capacity_Utilization)) +
    geom_line(colour = "darkorchid") +
    theme(panel.background = element_rect(fill = "#eeeeee"),
          text = element_text(size = 18)) +
    labs(y="Index",
         x="Year/Qtr",
         title="Capacity Utilization")
  
```

*** 
###### Real GDP Growth over Prior Quarter, Seasonally Adjusted Annualized Rate  
###### Real GDP Growth over Prior Year, Seasonally Adjusted Annualized Rate  
###### Real GDP Gap: 100 * (Real GDP - GDP Potential)/GDP Potential;  Source: Real Potential Gross Domestic Product - Billions of Chained 2012 Dollars - Not Seasonally Adjusted  
###### Capacity Utilization: Total Index - Percent of Capacity - Seasonally Adjusted	G.17 Industrial Production and Capacity Utilization, Board of Governors of the Federal Reserve System (US)	For a given industry, the capacity utilization rate is equal to an output index divided by a capacity index. The Federal Reserve Board's capacity indexes attempt to capture the concept of sustainable maximum output-the greatest level of output a plant can maintain within the framework of a realistic work schedule, after factoring in normal downtime and assuming sufficient availability of inputs to operate the capital in place.  
###### SOURCE: Employment Situation, U.S. Bureau of Labor Statistics
*** 

#### **The Unemployment Rate may be: 
- **overstated** by the **inclusion** of a **growing elderly population**, and 
- **understated** by the **exclusion** of **discouraged workers** in the working age population

```{r m_plots, figures-side, out.width="100%", fig.height=1.5, message=FALSE, echo=FALSE}

# Prep steps from facetsM.R
library(tidyverse)

df_m_raw <- read_tsv("data/Facets_Monthly_API.txt")
# colnames(df_m)

df_m <- df_m_raw[,!(colnames(df_m_raw) %in% c("PAYEMS", "USPRIV", "USGOVT",
                                            "CES9091000001", "CES9092000001",
                                            "CES9093000001"))]

# change series names to make them more readable
colnames(df_m) <- c("Yr_Mo", "Avg_Weekly_Hours", "Avg_Hrly_Earnings", "Population_over16", "CPI_All_Urban_All_Items", "CPI_All_Urban_All_Items_ex_Food_Energy", "01_Year_Treasury", "10_Year_Treasury", "05_Year_Treasury", "Working_Age_Pop_16to65", "Civilian_Labor_Force_Level", "Unemployment_Level", "Not_In_Labor_Force", "PCE", "PCE_ex_Food_Energy", "Trimmed_Mean_PCE_Inflation_Rate", "Population", "Price_Pressures_Measure")
# colnames(df_m)

# Drop all rows where ??? = NA
df_m <- df_m[is.na(df_m$Avg_Weekly_Hours)==FALSE,]

# transform data
df_m$Employment_Level <- df_m$Civilian_Labor_Force_Level - df_m$Unemployment_Level
df_m$Labor_Force_Participation_Rate <- 100 * (df_m$Civilian_Labor_Force_Level / df_m$Population_over16)
df_m$Unemployment_Rate <- 100 * df_m$Unemployment_Level / df_m$Civilian_Labor_Force_Level
df_m$Unemployed_PctPop16 <- 100 * df_m$Unemployment_Level /df_m$Population_over16
df_m$Employed_PctPop16 <- 100 * df_m$Employment_Level /df_m$Population_over16
df_m$NotInLF_PctPop16 <- 100 * (df_m$Not_In_Labor_Force / df_m$Population_over16)
df_m$Population_over65 <- df_m$Population_over16 - df_m$Working_Age_Pop_16to65
df_m$Population_under16 <- df_m$Population - df_m$Civilian_Labor_Force_Level
df_m$Population_under16_PctTotalPop <- 100 * (df_m$Population_under16 / df_m$Population)
df_m$Population_over65_PctTotalPop <- 100 * (df_m$Population_over65 / df_m$Population)
df_m$Population_16to65_PctTotalPop <- 100 - df_m$Population_under16_PctTotalPop - df_m$Population_over65_PctTotalPop

# subset population percentages
df_m_pct <- df_m[,c("Yr_Mo","Employed_PctPop16", "Unemployed_PctPop16",
                    "NotInLF_PctPop16")]

# Reshape dataframe to graph multiple lines
# need to convert multiple variable to single with multiple categories

df_pop_cat <- gather(df_m_pct, 
                     key = "Population", 
                     value = "Percent", 
                     -Yr_Mo,
                     na.rm=FALSE)

# Fix names in Population Column
df_pop_cat$Population <- gsub("Employed_PctPop16", 
                              "Employed", 
                              df_pop_cat$Population)

df_pop_cat$Population <- gsub("Unemployed_PctPop16", 
                              "Unemployed", 
                              df_pop_cat$Population)

df_pop_cat$Population <- gsub("NotInLF_PctPop16", 
                              "Not in Labor Force", 
                              df_pop_cat$Population)

# Make 'Population' an ordered factor
df_pop_cat$Population <- ordered(df_pop_cat$Population,
                                 levels = c("Employed",
                                            "Not in Labor Force",
                                            "Unemployed"))

# plot: KEEP THIS ONE!!!!!!! ####

ggplot(data = df_pop_cat,
       aes(x=Yr_Mo,
           color=Population,
           y=Percent)) +
  geom_line() +
  scale_colour_hue(l = 40,
                   h = c(160, 300) + 15,
                   c = 100) +
  theme(panel.background = element_rect(fill = "#eeeeee", colour = "grey"),
        text = element_text(size=6) ) +
  labs(y="Percent of Total") 

```

###### NOTE1: Monthly, Not Seasonally Adjusted 
###### NOTE2: Population = Under 16 + Labor Force (i.e., Employed + Unemployed) + Over 16 Not in Labor Force
###### SOURCE: Employment Situation, U.S. Bureau of Labor Statistics
*** 

<!-- #### **Population by Age** -->
<!-- | | | | | -->
<!-- |:-|:-|:-|:-| -->
<!-- | *Real GDP Growth over Prior Quarter SAAR* | *Real GDP Growth over Prior Year SAAR* | *Real GDP % of Potential (+over/-underperforming) QSAAR* | *Capacity Utilization, SA* | -->


<!-- #### **Employment** -->
```{r monthly_plots1, figures-side, out.width="33%", fig.height=5, message=FALSE, echo=FALSE}

# # m plots ####

  ggplot(data = df_m, aes(x = Yr_Mo, y = Unemployed_PctPop16)) +
    theme(panel.background = element_rect(fill = "#eeeeee"),
          text = element_text(size=20)) +
    labs(title="Unemployment",
         y="Percent", x="Monthly",
         caption="Source: Employment Situation, U.S. Bureau of Labor Statistics") +
    labs(subtitle="Percent of Labor Force") +
    geom_ribbon(aes(ymin=min(Unemployed_PctPop16)-1, ymax=Unemployed_PctPop16), fill="darkorchid")

  ggplot(data = df_m, aes(x = Yr_Mo, y = Employed_PctPop16)) +
    theme(panel.background = element_rect(fill = "#eeeeee")) +
    labs(title="Real GDP Growth over Prior Year Potential",
         y="Percent", x="Monthly",
         caption="Source: Employment Situation, U.S. Bureau of Labor Statistics") +
    labs(subtitle="Percent of Population Over 16") +
    geom_ribbon(aes(ymin=min(Employed_PctPop16)-1, ymax=Employed_PctPop16), fill="darkorchid")

  ggplot(data = df_m, aes(x = Yr_Mo, y = NotInLF_PctPop16)) +
    theme(panel.background = element_rect(fill = "#eeeeee")) +
    labs(title="Not in Labor Force",
         y="Percent", x="Monthly",
         caption="Source: Employment Situation, U.S. Bureau of Labor Statistics") +
    labs(subtitle="Percent of Population Over 16") + 
    geom_ribbon(aes(ymin=min(NotInLF_PctPop16)-1, ymax=NotInLF_PctPop16), fill="darkorchid")


```

###### NOTE: Population = Under 16 + Labor Force (i.e., Employed + Unemployed) + Over 16 Not in Labor Force
###### SOURCE: Employment Situation, U.S. Bureau of Labor Statistics
***

### **Employment**

Non-Farm employees (1000s of persons)
- Employment Situation Report - Seasonally Adjusted
- Source: U.S. Bureau of Labor Statistics 

```{r emp_plots, figures-side, out.width="70%", fig.height=2.75, message=FALSE, echo=FALSE}

# # NON-FARM EMPLOYEES ‘000s of persons: 
#   U.S. Bureau of Labor Statistics - Employment Situation Report - Seasonally Adjusted
#
#         "PAYEMS", # - Total
#         "USPRIV", # - Private
#         "USGOVT", # - Government
#         "CES9091000001", # - Federal
#         "CES9092000001", # - State
#         "CES9093000001" # - Local

df_emp <- df_m_raw[,(colnames(df_m_raw) %in% c("DATE",
                                               # "PAYEMS", 
                                               "USPRIV", "USGOVT"))]
df_emp_govt <- df_m_raw[,(colnames(df_m_raw) %in% c("DATE", 
                                                    # "USGOVT",
                                                    "CES9091000001","CES9092000001", "CES9093000001"))]


# change series names to make them more readable
colnames(df_emp) <- c("Yr_Mo", 
                      # "Total_Nonfarm_Employees", 
                      "Private_Employees", "Govt_Employees")

colnames(df_emp_govt) <- c("Yr_Mo", 
                           # "Govt_Employees",
                           "Federal_Employees", "State_Employees", "Local_Employees")

# Drop all rows where with NA
df_emp <- df_emp[is.na(df_emp$Private_Employees)==FALSE,]
df_emp_govt <- df_emp_govt[is.na(df_emp_govt$Local_Employees)==FALSE,]

# Reshape dataframe to graph multiple lines
# need to convert multiple variable to single with multiple categories

df_emp_cat <- gather(df_emp, 
                     key = "Total Employees",
                     value = "Count",
                     -Yr_Mo,
                     na.rm=FALSE)

df_emp_govt_cat <- gather(df_emp_govt,
                          key = "Gov Employees",
                          value = "Count",
                          -Yr_Mo,
                          na.rm=FALSE)

# Fix names in Population Column
df_emp_cat$`Total Employees` <- gsub("Total_Nonfarm_Employees", 
                              "Total", 
                              df_emp_cat$`Total Employees`)

df_emp_cat$`Total Employees` <- gsub("Private_Employees", 
                              "Private", 
                              df_emp_cat$`Total Employees`)

df_emp_cat$`Total Employees` <- gsub("Govt_Employees", 
                              "Government", 
                              df_emp_cat$`Total Employees`)

df_emp_govt_cat$`Gov Employees` <- gsub("Govt_Employees", 
                              "Government", 
                              df_emp_govt_cat$`Gov Employees`)

df_emp_govt_cat$`Gov Employees` <- gsub("Federal_Employees", 
                              "Federal", 
                              df_emp_govt_cat$`Gov Employees`)

df_emp_govt_cat$`Gov Employees` <- gsub("State_Employees", 
                              "State", 
                              df_emp_govt_cat$`Gov Employees`)

df_emp_govt_cat$`Gov Employees` <- gsub("Local_Employees", 
                              "Local", 
                              df_emp_govt_cat$`Gov Employees`)

# Make 'Population' an ordered factor
df_emp_cat$`Total Employees` <- ordered(df_emp_cat$`Total Employees`,
                                 levels = c("Total",
                                            "Private",
                                            "Government"
                                            ))

# Make 'Population' an ordered factor
df_emp_govt_cat$`Gov Employees` <- ordered(df_emp_govt_cat$`Gov Employees`,
                                           levels = c("Government",
                                            "Federal",
                                            "State",
                                            "Local"
                                            ))

# plot: ####
p_emp <- ggplotly(
  dynamicTicks = TRUE, 
  originalData = FALSE,
ggplot(data = df_emp_cat,
       aes(x=Yr_Mo,
           color=`Total Employees`,
           y=Count)) +
  geom_line() +
  scale_colour_hue(l = 40,
                   h = c(160, 300) + 15,
                   c = 100) +
  theme(panel.background = element_rect(fill = "#eeeeee", colour = "grey"),
        text = element_text(size=6) ) +
  labs(y="Thousands"))

p_emp_govt <- ggplotly(
  dynamicTicks = TRUE, 
  originalData = FALSE,
ggplot(data = df_emp_govt_cat,
       aes(x=Yr_Mo,
           color=`Gov Employees`,
           y=Count)) +
  geom_line() +
  scale_colour_hue(l = 40,
                   h = c(160, 300) + 15,
                   c = 100) +
  theme(panel.background = element_rect(fill = "#eeeeee", colour = "grey"),
        text = element_text(size=6) ) +
  labs(y="Thousands"))

p_emp

p_emp_govt

```

###### NOTE1: Monthly, Seasonally Adjusted 
###### SOURCE: Employment Situation, U.S. Bureau of Labor Statistics
*** 



